---
title: Algebraic maximum likelihood estimators
output:
  github_document:
    toc: true
#always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: `algebraic.mle`

<!-- badges: start -->
<!-- badges: end -->

An algebra over maximum likelihood estimators (MLE).

MLEs have many desirable, well-defined statistical properties.
We define an algebra over MLEs.

## Installation

You can install the development version of `algebraic.mle` from [GitHub](https://github.com/) with:

```{r,eval=F}
install.packages("devtools")
devtools::install_github("queelius/algebraic.mle")
```

## API

The object representing a fitted model is a type of `mle` object, the maximum
likelihood estimator of the model with respect to observed data.

For a full list of functions, see the [function reference](https://queelius.github.io/algebraic.mle/reference/index.html)
for `algebraic.mle`.
In what follows, we briefly define the API (generic functions, mostly) with
default implementations for objects that inherit from `mle`:

1. `vcov(mle)` returns the variance-covariance matrix of the model's
   parameter estimates.
   `mse(mle)` computes the asymptotic mean squared error of the `mle` object.
   `sd(mle)` computes the asymptotic standard deviation of the `mle` object.
   `confint(mle)` returns the asymptotic confidence intervals of the components
   of the parameter value estimated by the `mle` object.
   `point(mle)` returns the point that maximizes a likelihood of the model.

2. `sampler(mle)` maps to a function that may be used to sample from the
   sampling distribution of the `mle` object.

3. `mle_weighted(mles)` computes the *weighted* maximum likelihood estimate
   from independent `mle` objects in the list argument `mles`.
   
4. `fisher_info(mle)` returns the Fisher information matrix of the model's
   parameters.
   
5. `rmap(mle,g)` provides an approximation of the maximum likelihood estimator
   of $g(\theta)$ where `mle` is an estimator of $\theta$. The returned value
   is an `numerical_mle` estimator of $g(\theta)$. Note that `numerical_mle`
   inherits from `mle`.
   
6. `linear_transform(A,mle)`

   The normal distribution is closed under linear combinations and
   transformations. For instance, let $A$ be a $p$-by-$q$ matrix and
   $\hat\theta$ be a $q$ dimensional random vector (MLE of $\theta$), then
   $A\hat\theta$ is a $p$ dimensional multivariate normal with a mean $A\theta$
   and a variance-covariance
   $\operatorname{trans}(A)\operatorname{vcov}(\hat\theta)A$.
   By the invariance property of the MLE, $A\hat\theta$ is the MLE of $A\theta$.
   
8. `+(mle1,mle2)`, `-(mle1,mle2)`, `neg(mle)`, and `pos(mle)`
   
   Assuming `mle1` and `mle2` are independent `mle` objects, we also define
   addition and subtraction of `mle` objects, e.g.,
   $\hat\theta_1 + \hat\theta_2$ is asymptotically distributed as
   $$
     N(\mu = \theta_1+\theta_2,\sigma^2 = \sigma_{\hat\theta_1}^2 +
       \sigma_{\hat\theta_2}^2).
   $$
   This linear combination is the MLE of $\theta_1+\theta_2$.
