---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: `algebraic.mle`

<!-- badges: start -->
<!-- badges: end -->

An algebra over maximum likelihood estimators (MLE).

MLEs have many desirable, well-defined statistical properties.
We define an algebra over MLEs.

## Installation

You can install the development version of `algebraic.mle` from [GitHub](https://github.com/) with:

```{r,eval=F}
# install.packages("devtools")
devtools::install_github("queelius/algebraic.mle")
```

## MLE of rate parameter in exponential distribution
In what follows, to demonstrate the `algebraic.mle` R package, we consider a
random sample from exponentially distributed random variable
$X_i \sim \operatorname{EXP}(\lambda=1)$ for $i=1,\ldots,n$.

We generate an observation from this random sample with:
```{r,message=F,warning=F}
library(stats)
library(tidyverse)
n <- 1000
rate <- 1
x <- tibble(x=stats::rexp(n,rate))
```

We have observed a sample of size $n=`r n`$.
We show some observations from this sample (data frame) with:
```{r}
print(x)
```

We show a histogram of the sample, and a plot of the exponential function's
pdf, with:
```{r,message=F}
library(ggplot2)
ggplot(x, aes(x=x)) + geom_histogram(aes(y=..density..),alpha=.2) +
    xlim(0,6) + 
    geom_function(fun=dexp)
```

If we would like to estimate $\lambda$, we can do so using maximum likelihood
estimation as implemented by the `algebraic.mle` package:
```{r}
library(algebraic.mle)
rate.hat <- mle_exp(x$x)
summary(rate.hat)
```

We can show the point estimator with:
```{r}
point(rate.hat)
```

We can show the Fisher information and variance-covariance matrices with:
```{r}
fisher_info(rate.hat)
vcov(rate.hat)
```

(If `rate.hat` had been a vector, `vcov` would have output a variance-covariance
matrix. We may consider the above outputs $1 \times 1$ matrices.)

We can show the confidence interval with:
```{r}
confint(rate.hat)
```

## Sampling distribution of the MLE
In general, to estimate the sampling distribution, we generate $B=10000$ samples
(of size $`r n`$) and their corresponding estimators,
$\hat\theta^{(1)},\ldots,\hat\theta^{(B)}$.

Normally, we do not have $B$ samples, and if we did, we would gather
all $B$ samples into one sample (or used a weighted MLE), which would contain
more (Fisher) information about $\theta$.

However, a nice property of MLEs is that, asymptotically, they converge to
a normal distribution with a mean given by the true parameter, in this case
$\lambda$, and a variance-covariance given by the inverse of the Fisher
information matrix, i.e., $\hat\lambda \sim N(\lambda,I^{-1}(\lambda))$ where
$I$ is the Fisher information matrix, in this case $I(\lambda) = n/\lambda^2$.

We observe the empirical sampling distribution of $\hat\theta$ overlaid with
the theoretical asymptotic distribution with:
```{r}
B <- 1000
data0 <- numeric(length=B)
for (i in 1:B)
{
    x <- stats::rexp(n,rate)
    data0[i] <- point(mle_exp(x))
}
ggplot(tibble(rate.hat=data0), aes(x=rate.hat)) +
    geom_histogram(aes(y=..density..),alpha=.3,bins=50) +
    geom_function(fun=function(x) { dnorm(x,mean=rate,sd=rate/sqrt(n)) })
```

We do not know $\lambda$, but we may estimate it from a sample, and thus
we may approximate the sampling distribution of $\hat\lambda$ with
$N(\hat\lambda,I^{-1}(\hat\lambda))$.

Since we are only giving one sample, we cannot do as we did before to provide
$B$ estimates of $\lambda$.
However, we can sample from the approximation of the asymptotic distribution of
$\hat\lambda$ with:
```{r}
data1 <- sampler(rate.hat)(B)
```

We visually compare the two MLE samples, `data0` and `data1`, with:
```{r,message=F,warning=F}
data <- data.frame(values=c(data0,data1), group=c(rep("empirical",B),rep("approximation",B)))

ggplot(data,aes(x=values,fill=group)) +
    geom_histogram(aes(y=..density..),position="identity", alpha=0.2, bins=50)
```

Due to sampling error, we see that the approximation, the estimate of the
asymptotic sampling distribution $N(\hat\lambda,\hat\lambda/n^{1/2})$, is
shifted to the left of the sample from the true distribution, but they appear to
be quite similar otherwise.

## Invariance property of the MLE

An interesting property of an MLE $\hat\lambda$ is that the MLE of $g(\lambda)$
is given by $g(\hat\lambda)$.

The method `fn_distr` applied to `mle` objects takes a function of the `mle`
and a simulation sample size and then computes its MLE, i.e., returns an object
of type `fn_mle`, which is also an MLE with a an asymptotic sampling
distribution.

### Example
Suppose we are interested in $g(\lambda) = 2 \lambda$:
```{r}
g <- function(lambda) 2*lambda

g.hat <- fn_distr(rate.hat,g)
point(g.hat)
vcov(g.hat)
confint(g.hat)
```

Let's compare this with a simulation of the actual distribution.

We know that the MLE $\hat\lambda \sim N(\lambda,\lambda^2/n)$.
We seek a transformation $g(\hat\lambda)$ such taht its expectation is
$2 \lambda$, i.e., $g(\lambda) = 2\lambda$.
Then, $\operatorname{var}(g(\hat\lambda)) = 4 \lambda^2/n$.
Letting $\lambda=1$, we see that $g(\hat\lambda) \sim N(2 \lambda,4\lambda^2/n)$.
We simulate drawing $n=1000$ observations from 

```{r}
data4 <- rnorm(1000,2,2/sqrt(1000))
mean(data4)
var(data4)
```

## Weighted MLE: a weighted sum of maximum likelihood estimators

Since the variance-covariance of an MLE is inversely proportional to the
Fisher information that the MLE is defined with respect to, we can combine
multiple MLEs of $\theta$, each of which may be defined with respect to a
different kind of sample, to arrive at the MLE that incorporates the Fisher
information in all of those samples.

Consider $k$ mutually independent MLE estimators of parameter $\theta$,
$\hat\theta_1,\ldots,\hat\theta_k$, where $\hat\theta_j \sim N(\theta,I_j^{-1}(\theta))$.

Then, the maximum likelihood estimator of $\theta$ that incorporates all of
the data in $\hat\theta_1,\ldots,\hat\theta_k$ is given by the inverse-variance
weighted mean,
$$
    \hat\theta = \left(\sum I_j(\theta)\right)^{-1} \left(\sum I_j(\theta) \theta_j\right).
$$


To evaluate the performance of this estimator, we generate a sample of $N=1000$
from $\operatorname{EXP}(\lambda=1)$.
We compute the MLE for the observed sample, denoted by $\theta$.
We then divide the observed sample into $r=5$ sub-samples, each of size
$N/r=100$, and compute the MLE for each sub-sampled, denoted by
$\theta^{(1)},\ldots,\theta^{(r)}$.
Finally, we come these MLEs into the weighted MLE, denoted by $\theta_w$.

```{r}
N <- 1000
r <- 5
data3 <- rexp(N)
data3.mat <- matrix(data3,nrow=r)
mles <- list(length=r)
for (i in 1:r)
    mles[[i]] <- mle_exp(data3.mat[i,])

mle.wt <- mle_weighted(mles)
summary(mle.wt)

mle.tot <- mle_exp(data3)
summary(mle.tot)
```

We see that $\hat\theta$ and $\hat\theta_w$ model approximately the same sampling
distribution for $\theta$.
