---
output: github_document
#always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: `algebraic.mle`

<!-- badges: start -->
<!-- badges: end -->

An algebra over maximum likelihood estimators (MLE).

MLEs have many desirable, well-defined statistical properties.
We define an algebra over MLEs.

## Installation

You can install `algebraic.mle` from [GitHub](https://github.com/queelius/algebraic.mle) with:
```r
install.packages("devtools")
devtools::install_github("queelius/algebraic.mle")
```

## API

The object representing a fitted model is a type of `mle` object, the maximum
likelihood estimator of the model with respect to observed data.

The API mostly consists of generic methods with implementations for various
`mle` type objects.
For a full list of functions, see the [function reference](https://queelius.github.io/algebraic.mle/reference/index.html)
for `algebraic.mle`.

Here is a simple minimum working example (MWE) of fitting an exponential model
to exponential data.
```{r}
library(algebraic.mle)
# Fit an exponential model to exponential DGP (data generating process)
rate.hat <- mle_exp(rexp(100, 1))
summary(rate.hat)
```

You can see tutorials for more examples of using the package in the
[vignettes](https://queelius.github.io/algebraic.mle/articles/index.html).