---
title: Algebraic maximum likelihood estimators
output:
  github_document:
    toc: true
#always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: `algebraic.mle`

<!-- badges: start -->
<!-- badges: end -->

An algebra over maximum likelihood estimators (MLE).

MLEs have many desirable, well-defined statistical properties.
We define an algebra over MLEs.

## Installation

You can install the development version of `algebraic.mle` from [GitHub](https://github.com/) with:

```{r,eval=F}
install.packages("devtools")
devtools::install_github("queelius/algebraic.mle")
```

## API

The object representing a fitted model is a type of `mle` object, the maximum
likelihood estimator of the model with respect to observed data.

For a full list of functions, see the [function reference](https://queelius.github.io/algebraic.mle/reference/index.html)
for `algebraic.mle`.
In what follows, we briefly define the API (generic functions, mostly) with
default implementations for objects that inherit from `mle`:

1. `vcov(mle)` returns the variance-covariance matrix of the model's
   parameter estimates.

2. `point(mle)` returns the point that maximizes the likelihood
   of the model.

3. `confint(mle)` returns confidence intervals of the true parameter value
   estimated by the `mle` object.

4. `sampler(mle)` maps to a function that may be used to sample from the
   sampling distribution of the `mle` object.

5. `mse(mle)` computes the asymptotic mean squared error of the `mle` object.

6. `mle_weighted(mles)` computes the *weighted* maximum likelihood estimate
   from independent `mle` objects in the list argument `mles`.
   
7. `fisher_info(mle)` returns the Fisher information matrix of the model's
   parameters.
   
8. `rmap(mle,g)` provides an approximation of the maximum likelihood estimator
   of $g(\theta)$ where `mle` is an estimator of $\theta$. The returned value
   is an `numerical_mle` estimator of $g(\theta)$. Note that `numerical_mle`
   inherits from `mle`.
   
9. Finally, since normal distributions are closed under linear transformations,
   then letting $A$ be a $p$-by-$q$ matrix and $\hat\theta$ be a $q$
   dimensional random vector (MLE for estimating $\theta$), then $A\hat\theta$
   is a $p$ dimensional multivariate normal with a mean $A\theta$ and a
   variance-covariance $A'\operatorname{vcov}(\hat\theta)A$.
   Furthermore, by the invariance property of the MLE, $A\hat\theta$ is the
   MLE of $A\theta$.
   
   We define a set of generic functions for multiplying `mle` objects by
   (non-random) matrices and vectors to permit these sort of operations, with
   the asymptotic distributions of the operators applied to the `mle` objects
   exactly known if the asymptotic distribution of the `mle` objects are
   exactly known.
   
   See `rmap` for when the operation is non-linear.

