---
output: github_document
#always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

---
output: github_document
#always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: `algebraic.mle`

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://github.com/queelius/algebraic.mle/blob/master/LICENSE)
<!-- badges: end -->

`algebraic.mle` is an R package that provides an algebra over
Maximum Likelihood Estimators (MLEs). These estimators possess
many desirable, well-defined statistical properties which the package
helps you manipulate and utilize.

## Installation

`algebraic.mle` can be installed from GitHub by using the devtools package
in R:
``` r
install.packages("devtools")
devtools::install_github("queelius/algebraic.mle")
```

## Purpose

The primary focus of this package is the likelihood and the log-likelihood,
fundamental statistical concepts for a parametric model. They are integral
to both Bayesian and frequentist statistics, as well as for those who
prioritize likelihood. The algebraic.mle package enables easy handling of
MLEs, which, under certain conditions and assumptions (such as independence
and identical distribution (iid) for a sample), present numerous advantages,
including asymptotic normality and being the uniformly minimum variance
unbiased estimator of theta.

## API Overview

The main object in the `algebraic.mle` package is the `mle` object, which
represents a fitted model. The package provides a number of generic methods
designed for `mle` objects. A comprehensive list of functions is available
in the [function reference](https://queelius.github.io/algebraic.mle/reference/index.html)
for `algebraic.mle`.


## Example
Here is an example of fitting a conditional exponential model to some data
using `algebraic.mle`.
The true DGP is given by `Y | x ~ EXP(rate(x))` where
`rate(x) = exp(b0 + b1*x)`, and we do not care how `x` is distributed.

Let's fit a conditional exponential model to some data.
In this model, `Y | x ~ EXP(rate(x))` where `rate(x) = exp(b0 + b1*x)`.
First, let's define the DGP (data generating process):

```{r}
b0 <- -.1
b1 <- 0.5

dgp <- function(n, x) {
    rate <- exp(b0 + b1 * x)
    rexp(n, rate) + rnorm(n, 0, 1e-3)
}
```

Let's generate some date:
```{r}
n <- 75
set.seed(1231)
df <- data.frame(x = rep(NA, n), y = rep(NA, n))
for (i in 1:n) {
    x <- runif(1, -10, 10)
    y <- dgp(n = 1, x = x)
    df[i, ] <- c(x, y)
}
```


Now, we define two functions, `resp`, `rate`, and `loglik` function which will 
be used to define the model.
```{r}
resp <- function(df) df$y
rate <- function(df, beta) exp(beta[1] + beta[2] * df$x)
loglik <- function(df, resp, rate) {
  function(beta) sum(dexp(x = resp(df), rate = rate(df, beta), log = TRUE))
}
```

Let's fit the model. We'll use the `optim` function in `stats` to fit the model
and then wrap it into an `mle` object using `mle_numerical`.

```{r}
library(algebraic.mle)

# initial guess for the parameters
par0 <- c(0, 0)
names(par0) <- c("b0", "b1")

sol <- mle_numerical(optim(par = par0,
    fn = loglik(df, resp, rate),
    control = list(fnscale = -1),
    hessian = TRUE))
summary(sol)
```

Let's plot it:
```{r}
# plot the x-y points from the data frame
plot(df$x,df$y)

# now overlay a plot of the conditional mean
x <- seq(-10, 10, .1)
b0.hat <- point(sol)[1]
b1.hat <- point(sol)[2]
y.hat <- 1/exp(b0.hat + b1.hat*x)
y <- 1/exp(b0 + b1*x)
lines(x, y, col = "green", lwd = 10)
lines(x, y.hat, col = "blue", lwd = 10)
```


### Hypothesis test and model selection

Let's test the hypothesis that `b0 = 0` using a likelihood ratio test.
We can use the LRT because this null model is a special case (nested) of the
full model. The null model is `Y | x ~ EXP(rate(x))` where
`rate(x) = exp(b1*x)`, while the full model is `Y | x ~ EXP(rate(x))` where
`rate(x) = exp(b0 + b1*x)`.

```{r null, warning = FALSE, message = FALSE}
# construct null model where b1 = 0
rateb0_zero <- function(df, b1) exp(b1 * df$x)

# initial guess for the parameters
b1 <- 0
names(b1) <- "b1"

# fit the model under the null hypothesis
sol_null_b0 <- mle_numerical(optim(par = b1,
    fn = loglik(df, resp, rateb0_zero),
    control = list(fnscale = -1),
    hessian = TRUE,
    method = "BFGS"))
summary(sol_null_b0)
```

Let's compute the likelihood ratio test statistic and p-value:
```{r}
(lrt <- -2 * (loglike(sol_null_b0) - loglike(sol)))
pchisq(lrt, df = 1, lower.tail = FALSE) # compute the p-value
```

We see that the `p > 0.05`, so the data appears to be *compatible* with 
the null hypothesis `b0 = 0`.

If we wanted to do model selection, we could use the AIC:
```{r}
aic(sol)
aic(sol_null_b0)
```

Since the AIC of the null model is greater than the AIC of the full model, we
conclude that the data is "more" compatible with the full model.

So, even though the data appears to be compatible with the null hypothesis
according to the LRT, the AIC suggests that the full model is better.
We actually know the DGP and both models are reasonable approximations, but
neither of course models the DGP exactly.

"All models are wrong, but some are useful." - George Box

Eventually, if we have a sufficiently large sample, any model that is not
the DGP can be discarded, but reality is so complex that we will never have
a large enough sample and we will never be able to come up with a model that
is exactly the DGP.

Let's do another test, $b1 = 0$, i.e., it's an unconditional exponential model,
or just a standard exponential distribution.
```{r nullb1, warning = FALSE, message = FALSE}
rate_b1_zero <- function(df, b0) exp(b0)
# fit the model under the null hypothesis
sol_null_b1 <- mle_numerical(optim(par = 0,
    fn = loglik(df, resp, rate_b1_zero),
    control = list(fnscale = -1),
    hessian = TRUE,
    method = "BFGS"))
(lrt.null_b1 <- -2 * (loglike(sol_null_b1) - loglike(sol)))
pchisq(lrt.null_b1, df = 1, lower.tail = FALSE) # compute the p-value
```

This has a `p`-value of essentially zero, so we reject the null hypothesis
that `b1 = 0`.

You can see tutorials for more examples of using the package in the
[vignettes](https://queelius.github.io/algebraic.mle/articles/index.html).

