---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: `algebraic.mle`

<!-- badges: start -->
<!-- badges: end -->

An algebra over maximum likelihood estimators.

## Installation

You can install the development version of `algebraic.mle` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("queelius/algebraic.mle")
```

## Example: MLE of rate parameter in exponential distribution

Suppose we have a sample of $n=50$ draws from $\operatorname{EXP}(\lambda=1)$.
```{r}
n = 50
rate = 1
x <- stats::rexp(n,rate)
head(x)
```

Then, we can estimate $\lambda$ with:
```{r}
library(algebraic.mle)
(rate.hat <- mle_exp(x))
```

We can show the point estimator with:
```{r}
point(rate.hat)
```

We can show the confidence intervals with:
```{r}
mle_confint(rate.hat)
```

We can sample from the MLE $\hat\lambda$ with:
```{r}
library(ggplot2)
library(tibble)
rrate.mle <- sampler(rate.hat)
#df <- tibble(x=rrate.mle(100000))
#head(df)
```

## MLE from sum of MLEs

Consider $k$ mutually independent MLE estimators of parameter $\theta$,
$\hat\theta_1,\ldots,\hat\theta_k$, where $\hat\theta_j \sim N(\theta,I_j^{-1}(\theta))$.

Then, the maximum likelihood estimator of $\theta$ that incorporates all of
the data in $\hat\theta_1,\ldots,\hat\theta_k$ is given by the inverse-variance
weighted mean,
$$
    \hat\theta = \left(\sum I_j(\theta)\right)^{-1} \left(\sum I_j(\theta) \theta_j\right).
$$
