---
title: "Exponential distribution MLE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exponential distribution MLE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
To demonstrate the R package `algebraic.mle`, we consider the relatively
simple case of a random sample from i.i.d. exponentially distributed random
variables.

First, we load the R package `algebraic.mle` with:

```{r setup}
library(algebraic.mle)
```

# Generating a sample

We define the parameters of the i.i.d. random sample with:
```{r}
n <- 1000
rate <- 1
```

We generate a random sample $X_i \sim \operatorname{EXP}(\lambda=`r rate`)$ for
$i=1,\ldots,n$ with:
```{r,message=F,warning=F}
library(stats)
library(tidyverse)
x <- tibble(x=stats::rexp(n,rate))
```

We have observed a sample of size $n=`r n`$.
We show some observations from this sample (data frame) with:
```{r}
print(head(x,n=4))
```

We show a histogram of the sample overlaid with a plot of the exponential
function's pdf, with:
```{r,message=F,warning=F}
library(ggplot2)
ggplot(x, aes(x=x)) + geom_histogram(aes(y=..density..),alpha=.2) +
    xlim(0,6) + 
    geom_function(fun=dexp)
```

# Maximum likelihood estimation
If we would like to estimate $\lambda=`r rate`$, we can do so using maximum likelihood
estimation as implemented by the `algebraic.mle` package:
```{r}
rate.hat <- mle_exp(x$x)
summary(rate.hat)
```

We can show the point estimator with:
```{r}
point(rate.hat)
```

We can show the Fisher information and variance-covariance matrices with:
```{r}
fisher_info(rate.hat)
vcov(rate.hat)
```

(If `rate.hat` had been a vector, `vcov` would have output a variance-covariance
matrix. We may consider the above outputs $1 \times 1$ matrices.)

We can show the confidence interval with:
```{r}
confint(rate.hat)
```

## Sampling distribution of the MLE
In general, to estimate the sampling distribution, we generate $B=10000$ samples
(of size $`r n`$) and their corresponding estimators,
$\hat\theta^{(1)},\ldots,\hat\theta^{(B)}$.

Normally, we do not have $B$ samples, and if we did, we would gather
all $B$ samples into one sample (or used a weighted MLE), which would contain
more (Fisher) information about $\theta$.

However, a nice property of MLEs is that, asymptotically, they converge to
a normal distribution with a mean given by the true parameter, in this case
$\lambda$, and a variance-covariance given by the inverse of the Fisher
information matrix $I$, in this case $I(\lambda) = n/\lambda^2$, i.e.,
$\hat\lambda \sim N(\lambda,I^{-1}(\lambda))$.

We observe the empirical sampling distribution of $\hat\theta$ overlaid with
the theoretical asymptotic distribution with:
```{r}
B <- 1000
data0 <- numeric(length=B)
for (i in 1:B)
{
    x <- stats::rexp(n,rate)
    data0[i] <- point(mle_exp(x))
}
ggplot(tibble(rate.hat=data0), aes(x=rate.hat)) +
    geom_histogram(aes(y=..density..),alpha=.3,bins=50) +
    geom_function(fun=function(x) { dnorm(x,mean=rate,sd=rate/sqrt(n)) })
```

We do not know $\lambda$, but we may estimate it from a sample, and thus
we may approximate the sampling distribution of $\hat\lambda$ with
$N(\hat\lambda,I^{-1}(\hat\lambda))$.

Since we are only giving one sample, we cannot do as we did before to provide
$B$ estimates of $\lambda$.
However, we can sample from the approximation of the asymptotic distribution of
$\hat\lambda$ with:
```{r}
data1 <- sampler(rate.hat)(B)
```

We visually compare the two MLE samples, `data0` and `data1`, with:
```{r,message=F,warning=F}
data <- data.frame(values=c(data0,data1), group=c(rep("empirical",B),rep("approximation",B)))

ggplot(data,aes(x=values,fill=group)) +
    geom_histogram(aes(y=..density..),position="identity", alpha=0.2, bins=50)
```

Due to sampling error, we see that the approximation, the estimate of the
asymptotic sampling distribution $N(\hat\lambda,\hat\lambda/n^{1/2})$, is
shifted to the left of the sample from the true distribution, but they appear to
be quite similar otherwise.

## Invariance property of the MLE

An interesting property of an MLE $\hat\lambda$ is that the MLE of $f(\lambda)$
is given by $f(\hat\lambda)$.

The method `rmap` applied to an object `x` for which `is(x,"mle") == TRUE` and a
function `f` compatible with `point(x)` (and optionally a simulation
sample size) computes the MLE of `f(x)`.

### Example
We know that the MLE $\hat\lambda \sim N(\lambda,\lambda^2/n)$.
We seek a transformation $g(\hat\lambda)$ such that its expectation is
$2 \lambda$, i.e., $g(\lambda) = 2\lambda$:
```{r}
f <- function(lambda) 2*lambda
```

We compute the MLE of $2 \hat\lambda$ (using a simulation size $n=1000$) with:
```{r}
f.hat <- rmap(rate.hat,f,n=1000)
```

Note that the simulation size is only for estimating $f(\lambda)$; regardless
of how large large $n$ is, the MLE of $f(\hat \lambda)$ will have a sampling error
that is, at best, a function of the sampling error $\hat\lambda$.

Then, $\operatorname{var}(f(\hat\lambda)) = 4 \lambda^2/n$.
Letting $\lambda=1$, we see that, asymptotically,
$g(\hat\lambda) \sim N(2,4/n)$.

We expected a mean of $2$ and a variance of $4/n = 0.004$, which is pretty
close to what we obtained:
```{r}
point(f.hat)
vcov(f.hat)
```

# Weighted MLE: a weighted sum of maximum likelihood estimators

Since the variance-covariance of an MLE is inversely proportional to the
Fisher information that the MLE is defined with respect to, we can combine
multiple MLEs of $\theta$, each of which may be defined with respect to a
different kind of sample, to arrive at the MLE that incorporates the Fisher
information in all of those samples.

Consider $k$ mutually independent MLE estimators of parameter $\theta$,
$\hat\theta_1,\ldots,\hat\theta_k$, where $\hat\theta_j \sim N(\theta,I_j^{-1}(\theta))$.

Then, the maximum likelihood estimator of $\theta$ that incorporates all of
the data in $\hat\theta_1,\ldots,\hat\theta_k$ is given by the inverse-variance
weighted mean,
$$
    \hat\theta = \left(\sum I_j(\theta)\right)^{-1} \left(\sum I_j(\theta) \theta_j\right).
$$

## Example
To evaluate the performance of the weighted MLE, we generate a sample of
$N=1000$ observations from $\operatorname{EXP}(\lambda=1)$ and compute the MLE
for the observed sample, denoted by $\theta$.

We then divide the observed sample into $r=5$ sub-samples, each of size
$N/r=100$, and compute the MLE for each sub-sampled, denoted by
$\theta^{(1)},\ldots,\theta^{(r)}$.

Finally, we do a weighted combination these MLEs to form the weighted MLE,
denoted by $\theta_w$:
```{r}
N <- 1000
r <- 5
samp <- rexp(N)
samp.sub <- matrix(samp,nrow=r)
mles.sub <- list(length=r)
for (i in 1:r)
    mles.sub[[i]] <- mle_exp(samp.sub[i,])

mle.wt <- mle_weighted(mles.sub)

mle <- mle_exp(samp)
```

We show the results in the following table:
```{r}
df <- data.frame(
    c(point(mle.wt),point(mle)),
    c(vcov(mle.wt),vcov(mle)))

colnames(df) <- c("point estimate","variance")
rownames(df) <- c("weighted MLE", "MLE")
print(df)
```

We see that $\hat\theta$ and $\hat\theta_w$ model approximately the same sampling
distribution for $\theta$.

