---
title: "Fitting Common Distributions to a DGP"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Fitting Common Distributions to a DGP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{dplyr}
  %\VignetteDepends{tibble}
  %\VignetteDepends{algebraic.mle}
  %\VignetteDepends{boot}
  %\VignetteDepends{CDFt}
  %\VignetteKeyword{normal}
  %\VignetteKeyword{weibull}
  %\VignetteKeyword{DGP}
  %\VignetteKeyword{maximum likelihood estimation} 
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(digits=3)
options(scipen=999)
```


```{r setup, warning=F, message=F}
library(algebraic.mle)
library(algebraic.dist)
library(boot)
```

## Introduction
The goal of this vignette is to demonstrate using the R package `algebraic.mle`
for inference on maximum likelihood estimators (MLEs). We will simulate a dataset
whose true data generating process (DGP) is a mixture of Weibull and Normal
distributions. However, we will fit Weibull and Normal distributions separately to
the data to explore which provides a better fit.


## Data Simulation
First, here are the simulation parameters:
```{r sim}
n <- 100
err <- 0.1
shape <- 2
scale <- 10
theta = c(shape, scale)
set.seed(142334)
```

We simulate a sample of size $n = `r n`$ from the DGP:
$$
    T_i = W_i + \epsilon_i
$$
where
$$
    W_i \sim \operatorname{Weibull}(k = `r shape`, \lambda = `r scale`)
$$
and
$$
    \epsilon_i \sim \operatorname{normal}(\mu = 0, \sigma = `r err`).
$$

We can simulate a sample from this DGP mixture distribution using the
`rweibull` and `rnorm` functions:
```{r}
x <- rweibull(n = n, shape = shape, scale = scale) +
  rnorm(n = n, mean = 0, sd = err)
```

## Visualizing Data
Here are some observations:
```{r}
head(x, n = 4)
```

Visualizing the data is a good first step in the analysis of the data.
If the data is univariate or bivariate, we can plot a histogram of the data
pretty easily.
We show a histogram of the simulated data below:
```{r histo, fig.align='center', echo=F}
library(ggplot2)
ggplot(data.frame(x = x), aes(x = x)) +
    geom_histogram(color = "dark blue",
                   fill = "light blue",
                   bins = 25) +
    labs(title = "Simulated Data",
         subtitle = "Histogram")
```

## Parametrically Modeling the Data
If we only had this sample, what might we conclude?
This can be a very difficult problem.

If we were only interested in, say, *prediction*, and we had a sufficiently
large sample, we could use a non-parametric methods and "let the data speak for
itself." However, if we are interested in inference (e.g., explaining the
data) or the sample was small, then we usually need to make some assumptions
about the data.

In this case, we will assume that the data is drawn from a parametric
distribution. There are many well-known, named parametric distributions, e.g.,
Pareto, Weibull, and Normal, to name a few.
We will fit the Weibull and the Normal distributions, and compare the results.

## Maximum Likelihood Estimation
We will use maximum likelihood estimation (MLE) to estimate the parameters of
both the Weibull and the Normal, and then wrap these estimates into an `mle`
object provided by `algebraic.mle` package:
```{r}
fit_normal <- function(data) {
    loglik <- function(theta) {
        sum(dnorm(data, mean = theta[1], sd = sqrt(theta[2]), log = TRUE))
    }
    mu.hat <- mean(data)
    sigma2.hat <- mean((data - mu.hat)^2)
    H <- -numDeriv::hessian(loglik, c(mu.hat, sigma2.hat))
    mle(theta.hat = c(mu.hat, sigma2.hat),
        loglike = loglik(c(mu.hat, sigma2.hat)),
        score = numDeriv::grad(loglik, c(mu.hat, sigma2.hat)),
        sigma = MASS::ginv(H),
        info = H,
        obs = data,
        nobs = length(data),
        superclasses = c("mle_normal"))
}

fit_weibull <- function(data) {
    loglik <- function(theta) {
        sum(dweibull(data, shape = theta[1], scale = theta[2], log = TRUE))
    }

    sol <- stats::optim(
        par = c(shape, scale),
        fn = loglik,
        hessian = TRUE,
        method = "L-BFGS-B",
        lower = c(0, 0),
        #method = "Nelder-Mead",
        control = list(maxit = 10000, fnscale = -1))

    mle(theta.hat = sol$par,
        loglike = sol$value,
        sigma = MASS::ginv(-sol$hessian),
        info = -sol$hessian,
        obs = data,
        nobs = length(data),
        superclasses = c("mle_weibull"))
}

bias.mle_normal <- function(x, theta = NULL) {
    if (is.null(theta))
        theta <- params(x)
    c(0, -theta[2] / nobs(x))
}

theta.hat <- fit_normal(x)
summary(theta.hat)

theta.weibull <- fit_weibull(x)
summary(theta.weibull)
```

Let's plot the pdfs of the Weibull and normal distributions:
```{r fig.align='center', fig.height=4, fig.width=4, echo=F}
df1 <- data.frame(Value=rweibull(n=100000,
    shape=params(theta.weibull)[1],
    scale=params(theta.weibull)[2]), Group = "Weibull")

df2 <- data.frame(Value=rnorm(n=100000,
    mean=params(theta.hat)[1],
    sd=sqrt(params(theta.hat)[2])), Group = "Normal")

df3 <- data.frame(Value=rweibull(n = 100000,shape = shape, scale = scale) +
                        rnorm(n = 100000, mean = 0, sd = err), Group = "True")
df <- rbind(df1, df2, df3)
ggplot(df, aes(x = Value, fill = Group)) +
  geom_density(alpha = 0.3, adjust = 1.5) +
  scale_fill_manual(values = c(
    Weibull="red",
    Normal="green",
    True="purple")) +
  labs(title = "Density plot", x = "Value", y = "Density") +
  theme_minimal()
```

In purple, we have the true density (DGP). In red, we have the Weibull density.
In green, we have the normal density. From the plot, it's hard to tell which
distribution is a better fit to the DGP.

Interestingly, the tails of the true distribution seem a bit heavier than the
tails of the Weibull and Normal. This may suggest that a heavier-tailed model
may be a better fit, such as the lognormal distribution, but we will not
pursue this.

### Performance Measures of the MLE
A nice property of MLEs is that, asymptotically, given some regularity
conditions, they are normally distributed with a mean given by the true
true parameter and a variance-covariance given by the inverse of
the FIM evaluated at $\theta$.

We do not know $\theta$, but we have have estimates, and thus
we may approximate the sampling distribution of $\hat\theta$ with
$\mathcal{N}(\hat\theta,I^{-1}(\hat\theta))$.

Let $F$ denote the true distribution function such that $X_j \sim F$ for all
$j$.
Suppose we have some population parameter $\theta = t(F)$
and an estimator of $\theta$ given by $\hat\theta = s(\{X_1,\ldots,X_n\})$.
A reasonable requirement for an estimator $\hat\theta$ is that it converges to
the true parameter value $\theta$ as we collect more and more data.
In particular, we say that it is a consistent estimator of $\theta$ if
$\hat\theta$ converges in probability to $\theta$, denoted by
$\hat\theta \overset{p}{\mapsto} \theta$.

If the regularity conditions hold for the MLE, then $\hat\theta$ is
a consistent estimator of $\theta$. However, for finite sample sizes, the
estimator may be biased.
The bias of $\hat\theta$ with respect to $\theta$ is defined as
$$
    \operatorname{bias}(\hat\theta,\theta) = E(\hat\theta) - \theta,
$$
where $\operatorname{bias}(\hat\theta,\theta) = 0$ indicates that $\hat\theta$
is an *unbiased* estimator of $\theta$.

As a function of the true distribution $F$, the bias is unknown and is not a
statistic.
However, in the case of the normal, $\hat\mu$ is unbiased and, analytically, the
bias of $\hat\sigma^2$ is given by $-\frac{1}{n} \sigma^2$:
```{r}
bias(theta.hat,theta)
```
If $\sigma^2$ is not known, we may estimate it by using replacing $\hat\sigma^2$
instead:
```{r}
bias(theta.hat)
```
This is pretty far off from the true bias. This may be the first indication that
the DGP is far from being normal.

If we wanted to estimate the bias for the Weibull, we could bootstrap it or something else,
but we don't attempt to do that here.

The mean squared error (MSE) is another performance measure of an estimator.
It is given by
$$
    \operatorname{mse}(\hat\theta) = E\bigl\{(\hat\theta - \theta)^T(\hat\theta - \theta)\bigr\},
$$
Another way to compute the MSE is given by
$$
    \operatorname{mse}(\hat\theta) =
        \operatorname{trace}(\operatorname{cov}(\hat\theta) +
        \operatorname{bias}(\hat\theta)^T
        \operatorname{bias}(\hat\theta).
$$

Here's R code to compute the MSE of $\hat\theta$:
```{r}
round(mse(theta.hat), digits=3)
round(mse(theta.weibull), digits=3)  # true MSE
```

The normal distribution has significant MSE compared to the Weibull.

## Invariance Property of the MLE

An interesting property of an MLE $\hat\theta$ is that the MLE of $f(\theta)$
is given by $f(\hat\theta)$. What is the distribution of $f(\hat\theta)$?
Asymptotically, it is normally distributed with a mean given by $f(\theta)$ and
a variace-covariance given by the covariance of the sampling distribution of
$f(\hat\theta)$.

We provide two methods to compute the variance-covariance.

### Delta Method
If $f$ is differentiable, the variance-covariance is given by
$$
\operatorname{var}(f(\hat\theta)) = \operatorname{E}\bigl\{
    \bigl(f(\hat\theta) - f(\theta)\bigr)^2\bigr\} =
    \operatorname{E}\bigl\{J_f(\hat\theta) I(\hat\theta)^{-1} J_f(\hat\theta)^T\bigr\}.
$$
Here, $J_f(\hat\theta)$ is the Jacobian of $f$ evaluated at $\hat\theta$.

### Monte-Carlo Method
The delta method requires that $f$ be differentiable, but we may use the
Monte-carlo method to estimate the distribution of $f(\hat\theta)$ for any
function $f$.
We simply sample from the MLE of $\hat\theta$ and apply $f$ to its estimates
and take the covariance of the sample.

Next, we show how to compute the sampling distribution of $g(\hat\theta)$
for some function $g$ and some MLE $\hat\theta$ using both the delta and mc
methods.

### Example 1
For this example, we use the Weibull fit.
Let $g(\theta) = A \theta + b$ for some matrix $A$ and vector $b$. (This is a
simple linear transformation of $\theta$.) We can define $g$ in R with:
```{r}
A <- matrix(c(2,3),nrow=2)
b <- c(1,0)
g <- function(x) A %*% x + b
```

We compute the variance-covariance of the MLE of $g(\theta)$ 
using both methods:
```{r}
g.mc <- rmap(theta.weibull,g,n=100000L)
g.delta <- rmap(theta.weibull,g,method="delta")
round(vcov(g.mc), digits=3)
round(vcov(g.delta), digits=3)
```

They are pretty close.

## Weighted MLE: A Weighted Sum of MLEs

Since the variance-covariance of an MLE is inversely proportional to the
Fisher information that the MLE is defined with respect to, we can combine
multiple MLEs of $\theta$, each of which may be defined with respect to a
different kind of sample, to arrive at the MLE that incorporates the Fisher
information in all of those samples.

Consider $k$ mutually independent MLEs of parameter $\theta$,
$\hat\theta_1,\ldots,\hat\theta_k$, where $\hat\theta_j \sim N(\theta,I_j^{-1}(\theta))$.
Then, the sampling MLE of $\theta$ that incorporates all of the data in
$\hat\theta_1,\ldots,\hat\theta_k$ is given by the inverse-variance
weighted mean,
$$
    \hat\theta_w = \left(\sum_{j=1}^k I_j(\theta)\right)^{-1} \left(\sum_{j=1}^k I_j(\theta) \hat\theta_j\right),
$$
which, asymptotically, has an expected value of $\theta$ and a variance-covariance
of $\left(\sum_{j=1}^k I_j(\theta)\right)^{-1}$.

### Example 2
For this example, we use the normal fit.

To evaluate the performance of the weighted MLE, we generate a sample of
$N=1000$ observations from $\mathcal{N}(\theta)$ and compute the MLE
for the observed sample, denoted by $\hat\theta$.

We then divide the observed sample into $r=5$ sub-samples, each of size
$N/r=100$, and compute the MLE for each sub-sampled, denoted by
$\theta^{(1)},\ldots,\theta^{(r)}$.

Finally, we do a weighted combination these MLEs to form the weighted MLE,
denoted by $\theta_w$:
```{r}
N <- 500
r <- 5
samp <- rnorm(N, mean = theta[1], sd = sqrt(theta[2]))
samp.sub <- matrix(samp, nrow = r)
mles.sub <- list(length = r)
for (i in 1:r)
    mles.sub[[i]] <- fit_normal(samp.sub[i,])

mle.wt <- mle_weighted(mles.sub)
mle <- fit_normal(samp)
```

We show the results in the following R code.
First, we show the weighted MLE and its MSE:
```{r}
params(mle.wt)
round(mse(mle.wt), digits=3)
```
The MLE for the total sample and its MSE is:
```{r}
params(mle)
round(mse(mle), digits=3)
```

We see that $\hat\theta$ and $\hat\theta_w$ model approximately the same sampling
distribution.

## Bootstrapping the MLEs
Let's compare the earlier results that relied on the large sampling assumption
with the bootstrapped MLE using `mle_boot`.
First, `mle_boot` is just a wrapper for `boot` objects or 
objects like `boot`. Thus to use `mle_boot`, we first need
to call `boot` to bootstrap our MLE for the Weibull fit.

We just need to wrap it in a function that
takes the data as input and returns the MLE of the parameters and then
pass it to `mle_boot` constructor:
```{r}
theta.boot <- mle_boot(
    boot(data = x,
         statistic = function(x, i) params(fit_weibull(x[i])),
         R = 1000))
```

We already printed out the `theta.boot` object, which provided a lot of
information about it, but we can obtain specified  statistics from the Bootstrap
MLE using the standard interface in `algorithmic.mle`, e.g.:
```{r}
print(theta.boot)
bias(theta.boot)
```

We see that, for the most part, the results are similar to those obtained
using the large sampling assumption.


## Goodness-of-Fit
We are fitting a model to the data that does not precisely capture the
generative model $W$. So, how good of a fit is it?

We will conduct a goodness of fit test,
\begin{align}
  H_0 &: \text{the data is compatible with the Weibull distribution}\\
  H_A &: \text{the data is not compatible with the Weibull distribution}.
\end{align}

To perform this test, we will use the Cramer-von Mises test. This test is
based on the Cramer-von Mises statistic, which is a measure of the distance
between the empirical distribution function of the data and the distribution
function of the model. The Cramer-von Mises statistic is given by
$$
  \hat D_n^2 = \frac{1}{n}\sum_{i=1}^n \left(\hat F_n(x_i) - F(x_i)\right)^2
$$
where $\hat F_n$ is the empirical distribution function of the data and
$F$ is the distribution function of the model.

```{r}
cramer.test <- function(obs.dat,ref.dat)
{
  stat <- CDFt::CramerVonMisesTwoSamples(obs.dat,ref.dat)
  list(p.value=exp(-stat)/6.0,
       cramer.stat=stat,
       obs.size=length(obs.dat),
       ref.size=length(ref.dat))
}

wei.shape <- params(theta.weibull)[1]
wei.scale <- params(theta.weibull)[2]
ref.dat <- rweibull(1000000, shape = wei.shape, scale = wei.scale)
cramer.test(x, ref.dat)
```

Looking at the $p$-value, we see that the data is compatible with the
Weibull distribution.
Now, let's do the same for the normal distribution:
```{r}
norm.mu <- params(theta.hat)[1]
norm.var <- params(theta.hat)[2]
ref.dat <- rnorm(1000000, mean = norm.mu, sd = sqrt(norm.var))
cramer.test(x, ref.dat)
```

They are both compatible with the data. However, the Weibull distribution
has a larger $p$-value, which may suggest it is a better fit.
We also have the AIC measure of goodness of fit. The AIC is given by
$$
  \text{AIC} = -2\log L + 2k,
$$
where $L$ is the likelihood of the model and $k$ is the number of parameters
in the model. The AIC is a measure of the tradeoff between the goodness of
fit and the complexity of the model.
```{r}
aic(theta.weibull)
aic(theta.hat)
```
A lower AIC value indicates a better fit. Thus, according to this measure,
the Weibull distribution is the better fit.

## Prediction Intervals

Frequently, we are actually interested in predicting the outcome of the
random variable (or vector) that we are estimating the parameters of.

We observed a sample $\mathcal{D} = \{T_i\}_{i=1}^n$ where $T_i \sim N(\mu,\sigma^2)$,
$\theta = (\mu,\sigma^2)^T$ is not known. We compute the MLE of $\theta$,
which, asymptotically, is normally distributed with a mean $\theta$ and a
variance-covariance $I^{-1}(\theta)/n$.

We wish to model the uncertainty of a new observation, $\hat{T}_{n+1}|\mathcal{D}$.
We do so by considering both the uncertainty inherent to the Normal distribution and
the uncertainty of our estimate $\hat\theta$ of $\theta$.
In particular, we let $\hat{T}_{n+1}|\hat\theta \sim N(\hat\mu,\hat\sigma^2)$
and $\hat\theta \sim N(\theta,I^{-1}(\theta)/n)$ (the sampling distribution of the
MLE).
Then, the joint distribution of $\hat{T}_{n+1}$ and $\hat\theta$ has the pdf
given by
$$
    f(t,\theta) = f_{\hat{T}|\hat\theta}(t|\theta=(\mu,\sigma^2)) f_{\hat\theta}(\theta),
$$
and thus to find $f(t)$, we marginalize over $\theta$, obtaining
$$
    f(t) = \int_{-\infty}^\infty \int_{-\infty}^{\infty} f_{\hat{T}_{n+1},\hat\mu,\hat\sigma^2}(t,\mu,\sigma^2) d\mu d\sigma^2.
$$

Given the information in the sample, the uncertainty in the new observation
is characterized by the distribution
$$
    \hat{T}_{n+1} \sim f(t).
$$

It has greater variance than $T_{n+1}|\hat\theta$ because, as stated earlier, we do
not know $\theta$, we only have an uncertain estimate $\hat\theta$.

In `pred`, we compute the predictive interval (PI) of the
distribution of $\hat{T}_{n+1}$ using Monte Carlo simulation, where we replace
the integral with a sum over a large number of draws from the joint distribution
of $\hat{T}_{n+1}$ and $\hat\theta$ and then compute the empirical quantiles.

The function `pred` takes as arguments `x`, in this case an `mle` object, 
and a sampler for the distribution of the random variable of interest, in this
case `rweibull` (the sampler for the normal distribution). The sampler
must be compatible with the output of `point(x)`, whether that output be
a scalar or a vector.
Here is how we compute the PI for $\hat{T}_{n+1}$:
```{r}
pred(x=theta.hat, samp=function(n=1, theta) rnorm(n,theta[1],theta[2]))
```

In general, it will return a $p$-by-$3$ matrix, where $p$ is the dimension of
$T$ and the columns are the mean, lower quantile, and upper quantile of the
predictive distribution.

How does this compare to $T_{n+1}|\hat\theta$? We can compute the 95% quantile
interval for $T_{n+1}|\hat\theta$ using the `qnorm` function:
```{r}
mu <- params(theta.hat)[1]
sd <- sqrt(params(theta.hat)[2])
c(mean=mu,lower=qnorm(.025,mean=mu, sd=sd),upper=qnorm(.975,mean=mu, sd=sd))
```

We see that the 95% quantile interval for $T_{n+1}|\hat\theta$ is smaller
than $\hat{T}_{n+1}$, which is what we expected. After all, there is
uncertainty about the parameter value $\theta$.

## Conclusion
In this vignette, we demonstrated how to use the `algebraic.mle` package to
estimate the sampling distribution of the MLE using the large sampling assumption
and the Bootstrap method. The package provides various functions for obtaining
statistics of the MLE, allowing for a deeper understanding of the properties of
your estimator.

We showed how to fit Weibull and Normal distributions
to a simulated dataset whose true distribution, while known, does not have
a common name.

We have shown how to compare the two models using the Cramer-von Mises
test and the AIC measure of goodness of fit. We came to no definitive
conclusion about which model is better, but the Weibull distribution has
a larger $p$-value from the Cramer-von Mises test, and a lower AIC value,
which serves as some evidence that it is a better fit. We saw the true DGP
is visually different from both the Weibull and the normal distributions.
Notably, the DGP has longer tails than both, suggesting that an even better
fit may be a long-tail distribution like the log-normal or
the Pareto distribution.