---
title: "Fitting models to unknown DGPs"
author: "Alex Towell"
date: "2022-10-14"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Fitting models to unknown DGPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{tibble}
---

```{r, include=F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
We are interested in the generative process that gave rise to the data we observed.
In the real world, systems are quite complex, but we settle for simpler models
for analytical tractability. So, we usually assume:

1. The sample is i.i.d.

2. The observations may be reasonably modeled by some relatively simple process.

3. Other factors, like ambient temperature, are either negligible or are more or
   less constant in the sample.

Since this is a simulation, we know the underlying DGP (data generating process).
It's just
$$
    T_i = W_i + \epsilon_i
$$
where
$$
    W_i \sim \operatorname{weibull}(k,\lambda)
$$
and
$$
    \epsilon_i \sim \operatorname{normal}(0,\sigma).
$$

In the real world, we do not know the DGP. In this study, we will assume that
either $T_1,\ldots,T_n$ comes from Weibull or Normal. Clearly, the true
DGF is a bit more complicated but still very simple compared to a more realistic
DGP.

Then, the process of parametrically modeling the observed data may take the
following steps:

1. Visualize the data, e.g., plot a histogram of the data.

2. Guess which parametric distribution (for the components) might fit the
   observed data for the system lifetime.
   
3. Use a statistical test for goodness-of-fit.

4. Repeat steps 2 and 3 if the measure of goodness of fit is not satisfactory.

## Simulation parameters and generation
The simulation parameters are given by:
```{r setup}
library(tibble)
library(stats)

sim.n <- 27
sim.err.sd <- 0.05
sim.shape <- 20
sim.scale <- 3
sim.theta = c(sim.shape,sim.scale)
set.seed(142334)
```

We generate the data with the following R code:
```{r}
sim.df <- tibble(lifetime=
  rweibull(n=sim.n, shape=sim.shape, scale=sim.scale) +
  rnorm(n=sim.n, mean=0, sd=sim.err.sd))
```

A few elements from the sample are given by:
```{r, echo=F}
head(sim.df)
```

## Visualizing the data
Visualizing the data is a good first step in the analysis of the data.
If the data is univariate or bivariate, we can plot a histogram of the data
pretty easily (if it's multivariate, we can plot the marginal distributions
of the data).

We show a histogram of the simulated data below:
```{r histo, fig.align='center', echo=F}
library(ggplot2)
ggplot(sim.df,aes(x=lifetime)) +
    geom_histogram(color="dark blue",
                   fill="light blue",
                   bins=25) +
    labs(title="Simulated Lifetime Data",
         subtitle="Histogram")
```

## Parametrically modeling the data
If we only had this sample, what might we conclude?
This can be a very difficult problem.
In our case, we know that the simulated data is drawn from the distribution
$T_i = W_i + \epsilon_i$ where
$$
  W_i \sim \operatorname{weibull}(\lambda = `r sim.shape`,
                                k = `r sim.scale`)
$$
and
$$
  \epsilon_i \sim \operatorname{normal}(\mu=0,\sigma=`r sim.err.sd`).
$$
However, in real-world data sets, we do not know the distribution. So, let us
suppose that we do not know the true distribution of the data.

If we were only interested in, say, *prediction*, and we had a sufficiently
large sample, we could use a non-parametric methods and "let the data speak for
itself." However, if we are interested in inference (e.g., explaining the
data) or the sample was small, then we usually need to make some assumptions
about the data.

In this case, we will assume that the data is drawn from a parametric
distribution. There are many well-known, named parametric distributions, e.g.,
Pareto, Weibull, and Normal, to name a few.
From experience, it seems like the Weibull and the normal might be good fits
to the data. However, note that since the normal distribution permits negative
values to be realized, it may not be an appropriate choice. Still, since these
are only approximations anyway, this may not be a big deal.

## Maximum likelihood estimation
First, let us fit the Weibull distribution by choosing appropriate shape
$\lambda$ and scale $k$ parameters using the maximum likelihood estimator.

To find the MLE of $\theta = (\lambda,k)'$, we need the log-likelihood function,
which is given by the following R code:
```{r}
loglike <- function(theta) sum(dweibull(
  sim.df$lifetime, shape=theta[1], scale=theta[2], log=T))
```

An MLE is a point $(\hat k,\hat\lambda)'$ that is a maximum of the loglikelihood
function, `loglike`, over the support of the parameters. Typically, closed
solutions aren't possible, so we normally use some sort of iterative technique.
We don't go into the details here, but normally a local search method, like
Newton-Raphson (which finds the value that makes the gradient of the loglikelihood
function zero) is used. However, these local methods -- being local -- need a good
starting point. 

Here is an example of Newton-raphson code, which we will not use since it is not
very efficient but is easy to understand:
```{r eval=F}
# f is the function we want to find the root of
# Jf is the jocabian of f
# x0 is the starting point
newton_raphson <- function(f, df, x0)
{
  eta <- 1 # learning rate, not too large to avoid overshooting
           # not too small to avoid slow convergence
  eps <- 1e-3 # close enough to zero to stop  
  repeat
  {
    fx <- f(x0)        # new function value
    if (max(abs(fx)) < eps) break # f(x) is close enough to zero
    J <- Jf(x0)        # jacobian
    d <- solve(J,fx)   # newton-raphson direction (pointing uphill)
    x0 <- x0 + eta * d # newton-raphson update (going uphill)
  }
  x0
}
```

We use a more efficient algorithm to compute the log-likelihood
function for the Weibull distribution. The code is given below:
```{r}
library(algebraic.mle)
ll.wei <- weibull_shape_scale_loglike(sim.df$lifetime)
```

### Numerical considerations
In the `algebraic.mle` package, we provide a precise
and efficient local iterative algorithm, `mle_weibull_shape_scale`,
for finding the MLE for the Weibull distribution.

As a local search method, it needs a good starting point for the shape
parameter $k$ close to the MLE, otherwise it may fail to converge
to the MLE.

To find a good starting point, we use a global search
method, Simulated Annealing, implemented by `sim_anneal` function.
Here is the code for finding a good starting point:
```{r}
# find a good starting position
start <- sim_anneal(
  f=ll.wei,
  x0=sim.theta,
  options=list(
    t_init=100,
    t_end=1e-4,
    alpha=0.99,
    iter_per_temp=200,
    sup=function(theta) all(theta > 0),
    trace=TRUE))
k0 <- start$argmax[1]
cat("initial guess for k0 =",k0,"\n")
```

Let's take a look at some plots for the simulated annealing algorithm:
```{r, fig.align='center', fig.height=4, fig.width=6, echo=F}
loglikes <- apply(start$path,1,ll.wei)
par(mfrow=c(1,3))
plot(loglikes, type="l",
  main="log-likelihood vs iteration",
  xlab="iteration", ylab="log-likelihood")
plot(start$path[,1], start$path[,2],
  type="p", main="exploration of support",
  xlab="shape (k)", ylab="scale (lambda)")
plot(start$path[,1],
  main="shape (k) vs iteration",
  xlab="iteration", ylab="shape (k)")
abline(h=sim.shape, col="red")
```

In the first plot, we see the history of log-likelihood values as the
algorithm progresses. The second plot shows the path of the algorithm
as it explores the support of the parameters. The third plot shows the
shape parameter, $k$, as the algorithm progresses. The red line is the
true value of $k$.

With this starting point in hand, we find an MLE with:
```{r}
library(algebraic.mle)
mle.wei <- mle_weibull_shape_scale(sim.df$lifetime, k0=k0)
```

The function, `mle_weibull_shape_scale`, returns an `mle` object, which has an
API that provides a number of conventient methods, such as estimating the
variance-covariance matrix, confidence intervals, bias, and so on.
Here's the code to print a summary of the MLE:
```{r}
summary(mle.wei)
```

Let's do the same for the normal distribution. We will use the
`mle_normal_mu_var` function from the `algebraic.mle` package:
```{r}
mle.norm <- mle_normal_mu_var(sim.df$lifetime)
summary(mle.norm)
```

Let's plot the pdfs of the Weibull and normal distributions:
```{r fig.align='center', fig.height=4, fig.width=4, echo=F}
df1 <- data.frame(Value=rweibull(n=100000,
    shape=point(mle.wei)[1],
    scale=point(mle.wei)[2]), Group = "Weibull")

df2 <- data.frame(Value=rnorm(n=100000,
    mean=point(mle.norm)[1],
    sd=sqrt(point(mle.norm)[2])), Group = "Normal")

df3 <- data.frame(Value=rweibull(n=100000,shape=sim.shape,scale=sim.scale) +
                        rnorm(n=100000,mean=0,sd=sim.err.sd), Group = "True")
df <- rbind(df1, df2, df3)
ggplot(df, aes(x = Value, fill = Group)) +
  geom_density(alpha = 0.3, adjust = 1.5) +
  scale_fill_manual(values = c(
    Weibull="red",
    Normal="green",
    True="purple")) +
  labs(title = "Density plot", x = "Value", y = "Density") +
  theme_minimal()
```

In purple, we have the true density (DGP). In red, we have the Weibull density.
In green, we have the normal density. From the plot, it's hard to tell which
distribution is a better fit to the DGP.

Interestingly, the tails of the true distribution seem a bit heavier than the
tails of the Weibull and Normal. This may suggest that a heavier-tailed model
may be a better fit, such as the lognormal distribution, but we will not
pursue this.

How do we choose between the Weibull and Normal distributions? We will
discuss this in the next section.

## Goodness of fit
We are fitting a model to the data that does not precisely capture the
generative model $W$. So, how good of a fit is it?

We will conduct a goodness of fit test,
\begin{align}
  H_0 &: \text{the data is compatible with the Weibull distribution}\\
  H_A &: \text{the data is not compatible with the Weibull distribution}.
\end{align}

To perform this test, we will use the Cramer-von Mises test. This test is
based on the Cramer-von Mises statistic, which is a measure of the distance
between the empirical distribution function of the data and the distribution
function of the model. The Cramer-von Mises statistic is given by
$$
  \hat D_n^2 = \frac{1}{n}\sum_{i=1}^n \left(\hat F_n(x_i) - F(x_i)\right)^2
$$
where $\hat F_n$ is the empirical distribution function of the data and
$F$ is the distribution function of the model.

```{r}
cramer.test <- function(obs.dat,ref.dat)
{
  stat <- CDFt::CramerVonMisesTwoSamples(obs.dat,ref.dat)
  list(p.value=exp(-stat)/6.0,
       cramer.stat=stat,
       obs.size=length(obs.dat),
       ref.size=length(ref.dat))
}

wei.shape <- point(mle.wei)[1]
wei.scale <- point(mle.wei)[2]
ref.dat <- rweibull(1000000,shape=wei.shape,scale=wei.scale)
cramer.test(sim.df$lifetime,ref.dat)
```

Looking at the $p$-value, we see that the data is compatible with the
Weibull distribution.
Now, let's do the same for the normal distribution:
```{r}
norm.mu <- point(mle.norm)[1]
norm.var <- point(mle.norm)[2]
ref.dat <- rnorm(1000000,mean=norm.mu,sd=sqrt(norm.var))
cramer.test(sim.df$lifetime,ref.dat)
```

They are both compatible with the data. However, the Weibull distribution
has a larger $p$-value, which may suggest it is a better fit.
We also have the AIC measure of goodness of fit. The AIC is given by
$$
  \text{AIC} = -2\log L + 2k,
$$
where $L$ is the likelihood of the model and $k$ is the number of parameters
in the model. The AIC is a measure of the tradeoff between the goodness of
fit and the complexity of the model.
```{r}
aic(mle.wei)
aic(mle.norm)
```
A lower AIC value indicates a better fit. Thus, according to this measure,
the Weibull distribution is the better fit.

## Conclusion
In this post, we have shown how to fit a Weibull and Normal distributions
to a simulated dataset whose true distribution, while known, does not have
a common name.

We have shown how to compare the two models using the Cramer-von Mises
test and the AIC measure of goodness of fit.

We came to no definitive conclusion about which model is better, but
the Weibull distribution has a larger $p$-value from the Cramer-von Mises
test, and a lower AIC value, which serves as some evidence that it is
a better fit. We saw the true DGP is visually different from both the Weibull and the normal
distributions. Notably, the DGP has longer tails than both, suggesting that
an even better fit may be a long-tail distribution like the log-normal or
the Pareto distribution.